---
title: "MovieLens Capstone Report"
author: "chsmithiii"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Executive Summary

**Goal.** Build a movie recommendation system with the **MovieLens 10M** dataset and report **RMSE** on a sealed final holdout.

**Approach.**
1. Reproducible pipeline: auto-download data, construct `edx` and `final_holdout_test` per course spec.
2. Develop models on a dev split of `edx`:
   - Global-mean baseline
   - **Regularized effects** (movie, user, optional primary-genre)
   - **Matrix factorization** via `recosystem` with tuned hyperparameters
   - **Linear ensemble** of effects + MF (weights learned on dev set only)
3. Retrain best components on the **full `edx`** and evaluate **once** on `final_holdout_test`.

**Highlights.**
- Regularization combats popularity/user sparsity bias; MF captures latent structure.
- Simple ensemble reliably improves dev RMSE, then is **frozen** for final scoring (no leakage).

# Data Acquisition & Splits

```{r packages}
if (!require(tidyverse)) install.packages("tidyverse", repos="http://cran.us.r-project.org")
if (!require(caret))      install.packages("caret", repos="http://cran.us.r-project.org")
if (!require(recosystem)) install.packages("recosystem", repos="http://cran.us.r-project.org")
library(tidyverse); library(caret); library(recosystem); library(stringr)
```

```{r seed-helper}
seed_ml <- function(s) {
  if (getRversion() >= "3.6.0") set.seed(s, sample.kind = "Rounding") else set.seed(s)
}
DEV <- FALSE   # TRUE for fast draft runs; FALSE for full training
```

```{r download-and-construct}
# Auto-download & unzip to relative ./data
options(timeout = 600)
dir.create("data", showWarnings = FALSE)
zip_path <- file.path("data", "ml-10m.zip")
if (!file.exists(zip_path)) {
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", destfile = zip_path, mode = "wb")
}
if (!dir.exists(file.path("data", "ml-10M100K"))) unzip(zip_path, exdir = "data")

ratings_file <- file.path("data", "ml-10M100K", "ratings.dat")
movies_file  <- file.path("data", "ml-10M100K", "movies.dat")

ratings <- readr::read_lines(ratings_file) %>%
  stringr::str_split_fixed("::", 4) %>%
  as.data.frame(stringsAsFactors = FALSE) %>%
  setNames(c("userId","movieId","rating","timestamp")) %>%
  mutate(across(c(userId, movieId, timestamp), as.integer),
         rating = as.numeric(rating))

movies <- readr::read_lines(movies_file) %>%
  stringr::str_split_fixed("::", 3) %>%
  as.data.frame(stringsAsFactors = FALSE) %>%
  setNames(c("movieId","title","genres")) %>%
  mutate(movieId = as.integer(movieId))

movielens <- left_join(ratings, movies, by = "movieId")
rm(ratings, movies)

# Construct edx and final holdout (sealed) per spec
seed_ml(1)
idx <- createDataPartition(movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-idx, ]; temp <- movielens[idx, ]

final_holdout_test <- temp %>%
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")
removed <- anti_join(temp, final_holdout_test)
edx <- bind_rows(edx, removed)
rm(temp, removed, idx, movielens)
```

```{r dev-split}
# Development split from edx (for model selection & tuning only)
seed_ml(42)
dev_idx   <- createDataPartition(edx$rating, times = 1, p = 0.1, list = FALSE)
edx_train <- edx[-dev_idx, ]
edx_test  <- edx[ dev_idx, ] %>%
  semi_join(edx_train, by = "movieId") %>%
  semi_join(edx_train, by = "userId")
rm(dev_idx)
```

# Exploratory Analysis (brief)

```{r eda-summary}
summary(edx$rating)
tibble(n_users  = n_distinct(edx$userId),
       n_movies = n_distinct(edx$movieId))
```

```{r plot-rating-dist, fig.height=3.2}
edx %>% ggplot(aes(rating)) +
  geom_histogram(binwidth = 0.5, boundary = 0.5) +
  scale_x_continuous(breaks = seq(0.5, 5, 0.5)) +
  labs(title = "Rating Distribution (edx)", x = "Rating", y = "Count")
```

# Methods

## RMSE & Utility Helpers

```{r helpers}
RMSE <- function(y, yhat) sqrt(mean((y - yhat)^2))
clip_rating <- function(x) pmin(pmax(x, 0.5), 5)  # keep predictions within valid 0.5–5 range
extract_primary <- function(g) stringr::str_split_fixed(g, "\\|", 2)[,1]
```

## Baseline: Global Mean

```{r baseline}
mu_tr   <- mean(edx_train$rating)
rmse_mu <- RMSE(edx_test$rating, rep(mu_tr, nrow(edx_test)))
rmse_mu
```

## Regularized Effects (movie, user, optional primary genre)

**Rationale (math note).** Let \(r_{ui}\) be the rating by user \(u\) on movie \(i\) and \(\mu\) the global mean. Regularized effects estimate:

\[ b_i = \frac{\sum_{u}(r_{ui}-\mu)}{n_i + \lambda}, \quad
   b_u = \frac{\sum_{i}(r_{ui}-\mu-b_i)}{n_u + \lambda}. \]

Optionally a primary-genre effect \(b_g\) analogously. \(\lambda>0\) shrinks noisy estimates (e.g., very few ratings) towards 0, reducing variance.

```{r reg-effects-tune, cache=TRUE}
# Add a single-token primary genre to preserve 1-row-per-rating without exploding rows
edx_train <- edx_train %>% mutate(genre1 = extract_primary(genres))
edx_test  <- edx_test  %>% mutate(genre1 = extract_primary(genres))

lambda_grid <- c(1, 2, 3, 5, 7, 10, 15, 25)
score_lambda <- function(lambda){
  mu <- mean(edx_train$rating)
  b_i <- edx_train %>% group_by(movieId) %>% summarize(b_i = sum(rating - mu)/(n() + lambda), .groups="drop")
  b_u <- edx_train %>% left_join(b_i, by="movieId") %>% group_by(userId) %>%
    summarize(b_u = sum(rating - mu - b_i)/(n() + lambda), .groups="drop")
  b_g <- edx_train %>% left_join(b_i, by="movieId") %>% left_join(b_u, by="userId") %>%
    group_by(genre1) %>% summarize(b_g = sum(rating - mu - b_i - b_u)/(n() + lambda), .groups="drop")
  pred <- edx_test %>% left_join(b_i, by="movieId") %>% left_join(b_u, by="userId") %>% left_join(b_g, by="genre1") %>%
    transmute(pred = clip_rating(mu + coalesce(b_i,0) + coalesce(b_u,0) + coalesce(b_g,0))) %>% pull(pred)
  RMSE(edx_test$rating, pred)
}
rmse_by_lambda <- sapply(lambda_grid, score_lambda)
lambda_best <- lambda_grid[which.min(rmse_by_lambda)]
lambda_best; rmse_by_lambda
```

```{r reg-effects-dev-rmse}
# Fit dev effects at best lambda to report RMSE
mu <- mean(edx_train$rating)
b_i <- edx_train %>% group_by(movieId) %>% summarize(b_i = sum(rating - mu)/(n() + lambda_best), .groups="drop")
b_u <- edx_train %>% left_join(b_i, by="movieId") %>% group_by(userId) %>%
  summarize(b_u = sum(rating - mu - b_i)/(n() + lambda_best), .groups="drop")
b_g <- edx_train %>% left_join(b_i, by="movieId") %>% left_join(b_u, by="userId") %>% group_by(genre1) %>%
  summarize(b_g = sum(rating - mu - b_i - b_u)/(n() + lambda_best), .groups="drop")

pred_eff_dev <- edx_test %>% left_join(b_i, by="movieId") %>% left_join(b_u, by="userId") %>% left_join(b_g, by="genre1") %>%
  transmute(pred = clip_rating(mu + coalesce(b_i,0) + coalesce(b_u,0) + coalesce(b_g,0))) %>% pull(pred)
rmse_eff_dev <- RMSE(edx_test$rating, pred_eff_dev)
rmse_eff_dev
```

## Matrix Factorization (MF) via `recosystem`

**Concept (math note).** Factorize the sparse rating matrix \(R\in\mathbb{R}^{U\times I}\) into low-rank latent factors \(P\in\mathbb{R}^{U\times k}, Q\in\mathbb{R}^{I\times k}\) with regularization, minimizing \(\sum_{(u,i)\in\Omega} (r_{ui}-p_u^\top q_i)^2 + \lambda (\|p_u\|^2+\|q_i\|^2)\). This captures user/item embeddings beyond explicit biases.

```{r mf-dev, cache=TRUE}
# Prepare dev files for recosystem
seed_ml(7)
train_file <- tempfile(); test_file <- tempfile()
edx_train %>% select(userId, movieId, rating) %>% write.table(train_file, sep=" ", row.names=FALSE, col.names=FALSE)
edx_test  %>% select(userId, movieId)        %>% write.table(test_file,  sep=" ", row.names=FALSE, col.names=FALSE)

r_dev <- Reco()
if (DEV) {
  tune_opts <- list(dim = c(20), lrate = c(0.05), costp_l2 = c(0.1), costq_l2 = c(0.1),
                    nthread = max(1, parallel::detectCores()-1), niter = 10)
  final_niter <- 20
} else {
  tune_opts <- list(dim = c(20, 40, 60), lrate = c(0.05, 0.1), costp_l2 = c(0.01, 0.1, 0.5),
                    costq_l2 = c(0.01, 0.1, 0.5), nthread = max(1, parallel::detectCores()-1), niter = 40)
  final_niter <- 80
}
tuned <- r_dev$tune(data_file(train_file), opts = tune_opts)

# quiet logs in report PDF
invisible(capture.output(
  r_dev$train(data_file(train_file), opts = c(tuned$min, niter = final_niter,
                                              nthread = max(1, parallel::detectCores()-1)))
))

pred_file <- tempfile()
r_dev$predict(data_file(test_file), out_file(pred_file))
pred_mf_dev <- scan(pred_file)
pred_mf_dev <- clip_rating(pred_mf_dev)
rmse_mf_dev <- RMSE(edx_test$rating, pred_mf_dev)
rmse_mf_dev
```

## Linear Ensemble (effects + MF)

**Design.** Fit \(\hat{y} = w_1\,\hat{y}_{effects} + w_2\,\hat{y}_{MF}\) by OLS **on dev** (no intercept), then freeze \(w_1, w_2\) for final evaluation. To avoid degenerate negatives, clamp to nonnegative and renormalize.

```{r ensemble-dev}
fit <- lm(edx_test$rating ~ 0 + pred_eff_dev + pred_mf_dev)
w <- coef(fit)
w_eff <- unname(w["pred_eff_dev"]); w_mf <- unname(w["pred_mf_dev"])
# sanitize weights
w_eff <- ifelse(is.na(w_eff) || w_eff < 0, 0, w_eff)
w_mf  <- ifelse(is.na(w_mf)  || w_mf  < 0, 0, w_mf)
if (w_eff + w_mf == 0) { w_eff <- 0.5; w_mf <- 0.5 } else { s <- w_eff + w_mf; w_eff <- w_eff/s; w_mf <- w_mf/s }

pred_ens_dev <- clip_rating(w_eff*pred_eff_dev + w_mf*pred_mf_dev)
rmse_ens_dev <- RMSE(edx_test$rating, pred_ens_dev)
list(weights = c(effects = w_eff, mf = w_mf), rmse_ensemble = rmse_ens_dev)
```

# Results on Development Split

```{r results-table}
results <- tibble(
  Model = c("Global mean", 
            paste0("Regularized movie+user+genre (lambda=", lambda_best, ")"),
            "Matrix factorization (tuned)",
            "Ensemble (effects + MF)"),
  RMSE  = c(rmse_mu, rmse_eff_dev, rmse_mf_dev, rmse_ens_dev)
)
knitr::kable(results, digits=5, caption = "RMSE on edx_test (development)")
```

```{r tuned-params-table}
tuned_tbl <- tibble(
  dim = tuned$min$dim, lrate = tuned$min$lrate,
  costp_l2 = tuned$min$costp_l2, costq_l2 = tuned$min$costq_l2,
  niter = ifelse(DEV, 20, 80), w_eff = w_eff, w_mf = w_mf, lambda = lambda_best
)
knitr::kable(tuned_tbl, digits=5, caption = "Best MF params and ensemble weights")
```

```{r subgroup-check, cache=TRUE}
edx_test %>%
  mutate(pred = pred_ens_dev) %>%
  group_by(genre1) %>%
  summarize(RMSE = RMSE(rating, pred), n = n(), .groups="drop") %>%
  arrange(RMSE) %>%
  head(10) %>%
  knitr::kable(digits=4, caption="Top-10 genres by (lowest) dev RMSE")
```

# Final Model Training & Holdout Evaluation (no leakage)

```{r final-train-and-eval, cache=TRUE}
# Refit effects on FULL edx at lambda_best
edx <- edx %>% mutate(genre1 = extract_primary(genres))
mu_full <- mean(edx$rating)

b_i_full <- edx %>% group_by(movieId) %>% summarize(b_i = sum(rating - mu_full)/(n() + lambda_best), .groups="drop")
b_u_full <- edx %>% left_join(b_i_full, by="movieId") %>% group_by(userId) %>%
  summarize(b_u = sum(rating - mu_full - b_i)/(n() + lambda_best), .groups="drop")
b_g_full <- edx %>% left_join(b_i_full, by="movieId") %>% left_join(b_u_full, by="userId") %>% group_by(genre1) %>%
  summarize(b_g = sum(rating - mu_full - b_i - b_u)/(n() + lambda_best), .groups="drop")

predict_effects <- function(df){
  df %>% mutate(genre1 = extract_primary(genres)) %>%
    left_join(b_i_full, by="movieId") %>% left_join(b_u_full, by="userId") %>% left_join(b_g_full, by="genre1") %>%
    transmute(pred = clip_rating(mu_full + coalesce(b_i,0) + coalesce(b_u,0) + coalesce(b_g,0))) %>% pull(pred)
}

# Retrain MF on FULL edx (quiet logs)
full_train_file <- tempfile(); holdout_uim_file <- tempfile(); final_pred_file <- tempfile()
edx %>% select(userId, movieId, rating) %>% write.table(full_train_file, sep=" ", row.names=FALSE, col.names=FALSE)
final_holdout_test %>% select(userId, movieId) %>% write.table(holdout_uim_file, sep=" ", row.names=FALSE, col.names=FALSE)

r_full <- Reco()
invisible(capture.output(
  r_full$train(data_file(full_train_file), opts = c(tuned$min, niter = ifelse(DEV, 20, 80),
                                                    nthread = max(1, parallel::detectCores()-1)))
))
r_full$predict(data_file(holdout_uim_file), out_file(final_pred_file))
final_pred_mf <- scan(final_pred_file)
final_pred_mf <- clip_rating(final_pred_mf)

# Effects predictions on holdout
final_pred_eff <- predict_effects(final_holdout_test)

# Ensemble with frozen dev weights
final_pred <- clip_rating(w_eff*final_pred_eff + w_mf*final_pred_mf)

# Safety checks
stopifnot(length(final_pred_mf) == nrow(final_holdout_test))
stopifnot(length(final_pred)    == nrow(final_holdout_test))

final_rmse <- RMSE(final_holdout_test$rating, final_pred)
final_rmse

# Save predictions (optional for repo)
final_results <- final_holdout_test %>% select(userId, movieId, rating) %>% mutate(predicted = final_pred)
readr::write_csv(final_results, file.path("results_movielens_final_predictions.csv"))
```

# Conclusion

- **What worked:** Regularized effects + MF ensemble delivered the best dev RMSE and generalized to the holdout.
- **Why:** Effects model captures systematic biases (movies, users, coarse genre), while MF captures latent structures.
- **Trade-offs:** MF is more compute-intensive; effects are fast and interpretable. Ensemble balances both.
- **Limitations:** No temporal dynamics, implicit feedback, or item/content features beyond primary genre.
- **Future work:** Add time-aware biases, multi-genre embeddings, implicit feedback (clicks), calibrated uncertainty, or deep hybrid models.

# Reproducibility

```{r}
sessionInfo()
```

# References

- MovieLens 10M: https://grouplens.org/datasets/movielens/10m/
- `recosystem` (R): https://cran.r-project.org/package=recosystem
- HarvardX PH125.8x materials and project guidelines (course site)
